Review Paper Draft

### 2020-06-23 main.tex
Edits to 3.4 and 3.5

Despite the apparent dissimilarities between PC-Lasso and LMM-Lasso, it has been shown that the random effect formulation of model \ref{eqn:model} used by LMM-Lasso can be equivalently expressed in terms of fixed effects. \cite{zhang2015principal} derived this equivalence using a probabilistic PCA formulation, while \cite{hoffman2013correcting} used the singular value decomposition of $\bX$, and the decomposition of $\bK$ shown in equation \ref{eqn:k}.

Thus, whether modeled as a fixed or random effect, the underlying effect of $\bZ \bgamma$ on phenotype remains the same, but its treatment as a fixed or random effect has important implications in terms of model fitting and estimation. Including principal components as fixed effects requires estimation of a greater number of parameters, which may result in a loss of power \cite{zhang2015principal}. Additionally, determining the appropriate number of principal components to adjust for is non-trivial, and has been the subject of numerous studies \cite{patterson2006population, zhao2018practical}. It has been shown that including too many or too few principal components can result in power loss or increased Type I errors, respectively \cite{zhang2015principal}. However, in practice it is often recommended that the ten largest principal components be included \cite{zhao2018practical}. This is the implementation of PC-Lasso we will use throughout our simulations.

%------------------------------%
\subsection{Simulating environmental confounding}
Based on the LMM regression model \ref{eqn:lmm_reg}, numerous simulation studies of the performance of LMMs \cite{Rakitsch2012, bhatnagar2019simultaneous} simulate the quantity $\bX\bgamma$ as a random draw from a $N(\boldsymbol{0}, \sigmagg \bK)$. However, simulating environmental confounding according to this framework, where $\bgamma$ is a random quantity, "washes out" bias in $\hat{\bbeta}$ while inflating the variance. This is due to the fact that the sign of any element of the vector $\bX\bgamma$ will change from one simulation to the next. We can see this by considering the expectation and variance of low-dimensional OLS estimates of $\hat{\bbeta}$ based on model \ref{eqn:lmm_reg}:
\begin{align*}
    \mathbb{E}(\hat{\bbeta}) &= (\bX^T \bX)^{-1} \bX^T \mathbb{E}(\by)\\
    &=  (\bX^T \bX)^{-1} \bX^T \mathbb{E}(\bX\bbeta + \bX\bgamma + \beps)\\
    &\quad\\
    \mathbb{V}(\hat{\bbeta}) &= (\bX^T \bX)^{-1} \bX^T \mathbb{V}(\by)\\
    &=  (\bX^T \bX)^{-1} \bX^T \mathbb{V}(\bX\bbeta + \bX\bgamma + \beps)\\
    &=  (\bX^T \bX)^{-1} \bX^T \mathbb{V}(\bX\bgamma + \beps) \bX  (\bX^T \bX)^{-1}.
\end{align*}
Under the assumption that $\bgamma \sim N(\boldsymbol{0}, \frac{\sigmagg}{p} \bI_p)$, $\mathbb{E}(\hat{\bbeta}) = \bbeta, \mathbb{V}(\hat{\bbeta}) = \frac{\sigmagg}{p}\bI + \sigmaee (\bX^T \bX)^{-1}$. Considered in the context of our air quality example, such an assumption would be akin to subpopulation A being exposed to good air quality in one simulation, and bad air quality in the next. Instead, considering $\bgamma$ as a fixed quantity is more consistent with environmental heterogeneity and allows us to explore the bias in $\hat{\bbeta}$. Specifically, in the low-dimensional OLS setting, under the assumption that the quantity $\bgamma$ is fixed, $\mathbb{E}(\hat{\bbeta}) = \bbeta + \bgamma, \mathbb{V}(\hat{\bbeta}) = \sigmaee (\bX^T \bX)^{-1}$.

Although we derive these quantities under model \ref{eqn:lmm_reg}, where $\bZ = \bX$, as our assumed data-generating model, these results also hold for the more general model \ref{eqn:model}, where $\bZ \ne \bX$, by decomposing $\bZ \bgamma$ into components in and orthogonal to the column space of $\bX$. Assume $\exists \boldsymbol{\tau, \psi}: \bZ \bgamma = \bX \boldsymbol{\tau + \psi}$ where $\boldsymbol{P}_{\bX} = \bX (\bX^T \bX)^{-1} \bX^T$, $\boldsymbol{\tau} \in \mathcal{C}(\boldsymbol{P}_{\bX})$, and $\boldsymbol{\psi} \in \mathcal{N}(\boldsymbol{P}_{\bX})$. Then,
\begin{align*}
    \by &= \bX\bbeta + \bZ\bgamma + \beps\\
    &= \bX\bbeta + \bX \boldsymbol{\tau} + \boldsymbol{\psi} + \beps\\
    &=  \bX \bbeta' + \beps'
\end{align*}
where $\boldsymbol{\tau}$ quantifies bias in the estimated $\boldsymbol{\hat{\beta}'}$ and $\boldsymbol{\psi}$ is absorbed by the error term. Thus, given the nature of environmental confounding in our setting of interest, and in order to assess these methodological performance in terms of estimation accuracy, we simulate according to a fixed $\bgamma$ framework and using the following data-generating model
\begin{equation}
    \label{eqn:sim}
    \by = a \bX \bbeta + b \bZ \bgamma + c \beps,
\end{equation}
where $\by$ is a pseudophenotype, and $a, b, c$ are scaling factors determined such that the variance of $\by$ is equal to 1, and is partitioned into signal, environmental confounding, and random noise components,
\begin{equation}
    \label{eqn:parition_y}
    \mathbb{V}(\by) = \eta + (1 - \eta) \xi + (1-\eta)(1-\xi).
\end{equation}
Here, $\eta$ is the proportion of the variance of $\by$ attributable to genetic factors, commonly referred to as the narrow-sense heritability, and $\xi$ partitions the remaining variance of $\by$ into latent environmental heterogeneity and random noise. Given our modeling objective of accurately estimating the effect sizes of observed SNPs, $\eta$ also represents our data-generating model's signal to noise ratio (SNR).

We generated these pseudophenotypes using three types of high-dimensional ($p > n$) $\bX$ matrices: (1) simulated SNP data from independent subpopulations, (2) simulated SNP data from admixed subpopulations, and (3) real genome-wide SNP data from known caucasian ($n = 417$), African American ($n = 134$), and Hispanic ($n = 37$) subpopulations. \anna{cite Vanderwas data} For each $\bX$ matrix, $\bZ$ was a corresponding incidence matrix of 1s and 0s identifying subpopulation membership. 500 simulations were performed for each data type and parameter combination.

>>>>

\pb{I don't like calling these two methods ``equivalent''.  It immediately begs the question: if they're equivalent, why is one of the main aims of this paper to compare them?  I propose renaming this section ``Relationship between PC-Lasso and LMM-Lasso'' and then talking about both their similarities AND their differences.}

Despite the apparent dissimilarities between PC-Lasso and LMM-Lasso, it has been shown that the random effect formulation of model \ref{eqn:model} used by LMM-Lasso can be equivalently expressed in terms of fixed effects. \cite{zhang2015principal} derived this equivalence using a probabilistic PCA formulation, while \cite{hoffman2013correcting} used the singular value decomposition of $\bX$, and the decomposition of $\bK$ shown in equation \ref{eqn:k}. \pb{use \textbackslash eqref, not \textbackslash ref, for equations}

Thus, whether modeled as a fixed or random effect, the underlying effect of $\bZ \bgamma$ on phenotype remains the same, but its treatment as a fixed or random effect has important implications in terms of model fitting and estimation. Including principal components as fixed effects requires estimation of a greater number of parameters, which may result in a loss of power \cite{zhang2015principal}. Additionally, determining the appropriate number of principal components to adjust for is non-trivial, and has been the subject of numerous studies \cite{patterson2006population, zhao2018practical}. It has been shown that including too many or too few principal components can result in power loss or increased Type I errors, respectively \cite{zhang2015principal}. Nevertheless, a common practice is to include the ten largest principal components in the analysis \citep{zhao2018practical}, so this is the way we implement PC-Lasso throughout Section~\ref{sec:results}.

%------------------------------%
\subsection{Simulating environmental confounding}

In carrying out simulation studies of the LMM regression model \ref{eqn:lmm_reg}, a common practice \citep{Rakitsch2012, bhatnagar2019simultaneous} is to draw a new value of the quantity $\bX\bgamma$ from a $N(\boldsymbol{0}, \sigmagg \bK)$ for each replication of the analysis.  We argue that this is somewhat misleading from the perspective of understanding the statistical properties of the method. Averaging over situations in which the environmental confounding is constantly changing direction and magnitude ``washes out'' bias in $\hat{\bbeta}$ while inflating the variance. To see why this is the case, consider the OLS \pb{have to define OLS somewhere, if it's not already defined} estimates of $\hat{\bbeta}$ based on model \ref{eqn:lmm_reg}:
\begin{align*}
    \mathbb{E}(\hat{\bbeta}) &= (\bX^T \bX)^{-1} \bX^T \mathbb{E}(\by)\\
    &=  (\bX^T \bX)^{-1} \bX^T \mathbb{E}(\bX\bbeta + \bX\bgamma + \beps)\\
    &\quad\\
    \mathbb{V}(\hat{\bbeta}) &= (\bX^T \bX)^{-1} \bX^T \mathbb{V}(\by)\\
    &=  (\bX^T \bX)^{-1} \bX^T \mathbb{V}(\bX\bbeta + \bX\bgamma + \beps)\\
    &=  (\bX^T \bX)^{-1} \bX^T \mathbb{V}(\bX\bgamma + \beps) \bX  (\bX^T \bX)^{-1}.
\end{align*}
Under the assumption that $\bgamma \sim N(\boldsymbol{0}, \frac{\sigmagg}{p} \bI_p)$, $\mathbb{E}(\hat{\bbeta}) = \bbeta, \mathbb{V}(\hat{\bbeta}) = \frac{\sigmagg}{p}\bI + \sigmaee (\bX^T \bX)^{-1}$ \pb{this is not a sentence}. To return to our earlier example involving confounding due to air quality, randomly drawing a new value of $\bX\bgamma$ for each replication would mean that subpopulation A may be exposed to good air quality in one replication, and bad air quality in the next. Reporting the average across these situations is inconsistent with the question we are truly interested in, namely: what would happen if we took repeated samples from fixed populations A and B? Those repeated samples would show a systematic bias due to the difference in air quality, but this is masked if we generate new values of $\bX\bgamma$ with each replication. Considering $\bgamma$ as a fixed quantity is more consistent with the usual meaning of environmental confounding. Specifically, in the low-dimensional OLS setting, under the assumption that the quantity $\bgamma$ is fixed, $\mathbb{E}(\hat{\bbeta}) = \bbeta + \bgamma, \mathbb{V}(\hat{\bbeta}) = \sigmaee (\bX^T \bX)^{-1}$.

Although we derive these quantities under model \ref{eqn:lmm_reg}, where $\bZ = \bX$, these results also hold for the more general model \ref{eqn:model}, where $\bZ \ne \bX$, by decomposing $\bZ \bgamma$ into components in and orthogonal to the column space of $\bX$. Assume \pb{NO ASSUMPTION IS NECESSARY HERE!} $\exists \boldsymbol{\tau, \psi}: \bZ \bgamma = \bX \boldsymbol{\tau + \psi}$ where $\boldsymbol{P}_{\bX} = \bX (\bX^T \bX)^{-1} \bX^T$, $\boldsymbol{\tau} \in \mathcal{C}(\boldsymbol{P}_{\bX})$, and $\boldsymbol{\psi} \in \mathcal{N}(\boldsymbol{P}_{\bX})$ \pb{need to say projection matrix, null space, column space}. Then,
\begin{align*}
    \by &= \bX\bbeta + \bZ\bgamma + \beps\\
    &= \bX\bbeta + \bX \boldsymbol{\tau} + \boldsymbol{\psi} + \beps\\
    &=  \bX \bbeta' + \beps'
\end{align*}
where $\boldsymbol{\tau}$ quantifies bias in the estimated $\boldsymbol{\hat{\beta}'}$ and $\boldsymbol{\psi}$ is absorbed by the error term. In other words, any model specified in terms of $\bZ$ and $\bgamma$ is equivalent to a parameterization using $\bX$ and $\boldsymbol{tau}$.

\pb{should this paragraph move to next section?}
Thus, given the nature of environmental confounding in our setting of interest, and in order to assess these methodological performance in terms of estimation accuracy, we simulate according to a fixed $\bgamma$ framework and using the following data-generating model
\begin{equation}
    \label{eqn:sim}
    \by = a \bX \bbeta + b \bZ \bgamma + c \beps,
\end{equation}
where $\by$ represents the phenotype, and $a, b, c$ are scaling factors determined such that the variance of $\by$ is equal to 1, and is partitioned into signal, environmental confounding, and random noise components,
\begin{equation}
    \label{eqn:parition_y}
    \mathbb{V}(\by) = \eta + (1 - \eta) \xi + (1-\eta)(1-\xi).
\end{equation}
Here, $\eta$ is the proportion of the variance of $\by$ attributable to genetic factors, commonly referred to as the narrow-sense heritability, and $\xi$ partitions the remaining variance of $\by$ into latent environmental heterogeneity and random noise. \pb{So, for example, $\eta=0.5$ and $xi=0.7$ means that\ldots} Given our modeling objective of accurately estimating the effect sizes of observed SNPs, $\eta$ also represents our data-generating model's signal to noise ratio (SNR).

We generated these pseudophenotypes using three types of high-dimensional ($p > n$) $\bX$ matrices: (1) simulated SNP data from independent subpopulations, (2) simulated SNP data from admixed subpopulations, and (3) real genome-wide SNP data from known caucasian ($n = 417$), African American ($n = 134$), and Hispanic ($n = 37$) subpopulations. \anna{cite Vanderwas data} For each $\bX$ matrix, $\bZ$ was a corresponding incidence matrix of 1s and 0s identifying subpopulation membership \pb{Is this true for the linear mixing case as well?}. 500 simulations were performed for each data type and parameter combination.
